\LARGE \bf Planning with Attitude

### 2020-10-13 paper/iros_abstract.tex
IROS submission
Update IROS abstract

@@

    Many useful robotic systems---including quadrotors, airplanes, satellites, autonomous
    underwater vehicles, and quadrupeds---can perform arbitrarily large three-dimensional
    translations and rotations as part of their normal operation. While simply
    representing the attitude of these sytems is nontrivial, generating and tracking
    motion plans for floating-base systems is an even more challenging problem.

@@

    To address this lack of solver support, we formulate a straightforward, generic method for adapting
    existing Newton and gradient-based algorithms to correctly account for the group structure of
    rotations. Based entirely on basic mathematical tools from linear algebra and
    calculus, our method is computationally efficient and should be both accessible and
    familiar to most roboticists. We apply this method to the ALTRO solver
    \cite{howell2019altro} and demonstrate its performance on several constrained trajectory
    optimization problems.

@@

    We represent---following the Hamilton convention---a quaternion $\q \in \mathbb{H}$ as 
    a standard vector $q \in \R^4 := [q_s \;\; q_v^T]^T$ where $q_s \in \R$ and 
    $q_v \in \R^3$ are the scalar and vector part of the quaternion, respectively. We denote
    the composition of two quaternions as $q_3 = q_2 \otimes q_1$.

    \begin{figure}
        \centering
        \includegraphics[width=\columnwidth]{figures/quadflip_new.png}
        \label{fig:quadflip}
        \caption{Snapshots of quadrotor flip trajectory solved using a modified version of 
        ALTRO that correctly accounts for the group structure of rotations. These modifications
        make it easier to find trajectories that undergo large changes in attitude, such as
        the loop-de-loop shown here.}
    \end{figure}

@@

    The key idea of this section is that differential quaternions live in
    three-dimensional space instead of the four-dimensional space of quaternions; this
    idea should match our intuition given rotations are inherently three-dimensional and
    differential rotations should live in the same space as angular velocity, i.e.
    $\R^3$.

@@

        Let $\phi \in \R^3$ be a differential rotation applied to a function with
        unit quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(q \otimes \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by
        differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at
        $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} \begin{bmatrix} 
                            -q_v^T \\ 
                            sI_3 + \skewmat{q_v}
                        \end{bmatrix}
                        := \pdv{h}{q} G(q) 
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is referred to as the \textit{attitude Jacobian}, which
        essentially becomes a ``conversion factor'' allowing us to apply results from
        standard vector calculus to the space of unit quaternions. This form is
        particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be
        obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the
        same (up to a constant scaling factor) for any choice of three-parameter attitude
        representation.

@@

    These adjustments effectively modify the optimization so that it is performed on the
    quaternion error state (or Rodrigues Parameters) instead of the space of unit
    quaternions. In practice, this is analagous to the Multiplicative Extended Kalman
    Filter \cite{markley2014fundamentals} in the state estimation community.

    The algorithm only needs one small modification to the ``forward pass,'' where the
    trajectory is modified by simulating the system forward using the optimal feedback
    policy, $\delta u = K \delta x + d$, computed during the ``backward'' pass. Instead of 
    computing $\delta x$ using simple subtraction, we now compute the error state using the
    inverse Cayley map \eqref{eq:invcayley} for the quaternion states. The rest of the
    forward pass is unmodified. 
    %\todo{I think you can safely omit some of these details and just refer to the ALTRO paper.}

    \subsection{Examples}
        To demonstrate the effectiveness of the proposed approach, we tested the modified
        version of ALTRO by solving for a quadrotor flip trajectory (see Figure \ref{fig:quadflip}).
        Four soft ``waypoint'' costs were placed along the trajectory, encouraging the
        quadrotor to follow a loop. The trajectory was initialized with hover controls and a
        state trajectory that smoothly rotated the quadrotor one full rotation about the 
        x-axis while moving to the goal. 
        The modified version of ALTRO solved the problem in 270 ms and 26 iterations, while
        the original version that naively treated the quaternion as a vector in $\R^4$ 
        solved in 870 ms and 91 iterations. The original version failed to provide a nice,
        smooth loop and instead reversed direction halfway through the loop.

        It's also worthwhile to note that this problem cannot be solved as-is with any
        three-parameter representation because of the singularities that occur at 90, 180,
        and 360 degrees for Euler angles, Rodrigues parameters, and MRP's, respectively.
        A classic approach to overcome this issue is to do discrete switching of the 
        reference orientation, which requires previous knowledge of the topology of the
        optimal solution. 
        % \begin{figure}
        %     \centering
        %     \includegraphics[height=4cm, width=8cm]{figures/timing_chart.tikz}
        %     \caption{Timing results for the airplane barrell roll, quadrotor flip, and
        %     quadrotor zig-zag examples using iLQR with roll-pitch-yaw Euler angles (RPY),
        %     quaternions, and the current method. The timing result for the Quadrotor flip
        %     with Euler angles is omitted since it failed to converge.
        %     }
        %     \label{fig:timing_chart}
        % \end{figure}

@@

    We have proposed a general, accessible, and computationally efficient approach to
    doing planning and control for systems with one or more unit quaternions in their
    state vectors. The approach allows for straightforward adaptation of many
    gradient-based methods for optimal control and motion planning, which we demonstrated
    on the ALTRO solver. 
    Initial results suggest that our approach can significantly reduce solve times by 
    improving convergence rates for problems that require large changes in attitude over
    the course of the trajectory. Future work will extend these ideas to Newton-based methods
    such as direct collocation and other domains outside of trajectory optimization.

>>>>

% Planning and controlling trajectories for floating-base robotic systems that
% experience large attitude changes is challenging due to the nontrivial group
% structure of 3D rotations. This paper introduces a powerful, computationally
% efficient, and accessible approach for adapting existing Newton and
% gradient-based methods to correctly account for the group structure of
% rotations. We demonstrate the effectiveness of the approach by modifying the
% ALTRO solver to optimize a loop-de-loop trajectory for a quadrotor, enabling
% the solver to find a dramatically better solution in less than a third the
% time.

@@

    Many robotic systems---including quadrotors, airplanes, satellites, 
    underwater vehicles, and quadrupeds---can perform arbitrarily large three-dimensional
    translations and rotations as part of their normal operation. While simply
    representing the attitude of these sytems is nontrivial, generating and tracking
    motion plans for floating-base systems is an even more challenging problem.

@@

    To address this lack of solver support, we formulate a straightforward, generic
    method for adapting existing Newton and gradient-based algorithms to correctly
    account for the group structure of rotations. Based entirely on basic mathematical
    tools from linear algebra and calculus, our method is computationally efficient and
    should be both accessible and familiar to most roboticists. We apply this method to
    the ALTRO solver \cite{howell2019altro} and demonstrate its performance on a 
    constrained trajectory optimization problem.

@@

    We represent---following the Hamilton convention---a quaternion $\q \in \mathbb{H}$
    as a standard vector $q \in \R^4 := [q_s \;\; q_v^T]^T$ where $q_s \in \R$ and $q_v
    \in \R^3$ are the scalar and vector part of the quaternion, respectively. We denote
    the composition of two quaternions as $\q_3 = \q_2 \otimes \q_1$. We use
    $\skewmat{x}$ to denote the skew-symmetric matrix such that $\skewmat{x} y = x \times
    y$.

    \begin{figure}
        \centering
        \includegraphics[width=\columnwidth]{figures/quadflip.png}
        \label{fig:quadflip}
        \caption{Snapshots of quadrotor flip trajectory solved using a modified version
        of ALTRO that correctly accounts for the group structure of rotations. These
        modifications make it easier to find trajectories that undergo large changes in
        attitude, such as the loop-de-loop shown here.}
    \end{figure}

@@

    The key idea of this section is that differential quaternions live in
    three-dimensional space instead of the four-dimensional space of quaternions; this
    idea should match our intuition given rotations are inherently three-dimensional and
    differential rotations should live in the same space as angular velocities, i.e.
    $\R^3$.

@@

        Let $\phi \in \R^3$ be a differential rotation applied to a function with
        unit quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(q \otimes \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by
        differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at
        $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} \begin{bmatrix} 
                            -q_v^T \\ 
                            q_s I_3 + \skewmat{q_v}
                        \end{bmatrix}
                        := \pdv{h}{q} G(q) 
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is referred to as the \textit{attitude Jacobian}, which
        essentially becomes a ``conversion factor'' allowing us to apply results from
        standard vector calculus to the space of unit quaternions. This form is
        particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be
        obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the
        same (up to a constant scaling factor) for any choice of three-parameter attitude
        representation.

@@

    These adjustments effectively modify the optimization so that it is performed on 
    differential rotations (or Rodrigues Parameters) instead of the space of unit
    quaternions. In practice, this is analagous to the Multiplicative Extended Kalman
    Filter \cite{markley2014fundamentals} in the state estimation community.

    The iLQR algorithm derives a locally-optimal feedback policy of the form $\delta u =
    K \delta x + d$ during the ``backward pass'' of each iteration, which is then used to
    simulate the system forward during the ``forward pass''. Instead of computing $\delta
    x$ using simple substraction as we would in the original algorithm, we now compute the
    error using the inverse Cayley map \eqref{eq:invcayley} for the quaternion states. The
    rest of the forward pass is unmodified.

    \subsection{Examples}
        To demonstrate the effectiveness of the proposed approach, we tested the modified
        version of ALTRO by solving for a quadrotor flip trajectory (see Figure \ref{fig:quadflip}).
        Four soft ``waypoint'' costs were placed along the trajectory, encouraging the
        quadrotor to follow a loop. The trajectory was initialized with hover controls and a
        state trajectory that smoothly rotated the quadrotor one full rotation about the 
        x-axis while moving to the goal. 
        The modified version of ALTRO solved the problem in 140 ms and 22 iterations, while
        the original version that na\"ively treated the quaternion as a vector in $\R^4$ 
        solved in 420 ms and 58 iterations. The original version failed to provide a nice,
        smooth loop and instead reversed direction halfway through the loop.

        It's also worthwhile to note that this problem cannot be solved as-is with any
        three-parameter representation because of the singularities that occur at 90, 180,
        and 360 degrees for Euler angles, Rodrigues parameters, and MRP's, respectively.
        % A classic approach to overcome this issue is to do discrete switching of the 
        % reference orientation, which requires previous knowledge of the topology of the
        % optimal solution. 
        % \begin{figure}
        %     \centering
        %     \includegraphics[height=4cm, width=8cm]{figures/timing_chart.tikz}
        %     \caption{Timing results for the airplane barrell roll, quadrotor flip, and
        %     quadrotor zig-zag examples using iLQR with roll-pitch-yaw Euler angles (RPY),
        %     quaternions, and the current method. The timing result for the Quadrotor flip
        %     with Euler angles is omitted since it failed to converge.
        %     }
        %     \label{fig:timing_chart}
        % \end{figure}

@@

    We have proposed a general, accessible, and computationally efficient approach to
    doing planning and control for systems with one or more unit quaternions in their
    state vectors. The approach allows for straightforward adaptation of many
    gradient and Newton-based methods for optimal control and motion planning, which we demonstrated
    on the ALTRO solver. By exploiting the unique structure of both the trajectory
    optimization problem and the rotation group, the newly modified version of ALTRO will
    likely be able to solve more challenging problems with performance.
    Future work will extend these ideas to Newton-based methods such as direct
    collocation and other domains outside of trajectory optimization.
